Top-down and bottom-up cohesiveness\\in microbial community coalescence

### 2021-06-02 draft/draft_v4.tex
draft v4 added, need to polish a few details with Alvaro

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{upgreek} % upright greek letters for math
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{makecell} % more table customization
\usepackage{mathtools} % nice matrices
\usepackage{units} % nice fractions
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\zxrsetup{toltxlabel}
\zexternaldocument*{supp}
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[table,xcdraw]{xcolor} % text and table background colors
\usepackage[margin=2.5cm]{geometry} % margins
\usepackage{lineno} % line numbers
\usepackage{textcomp} % special characters (some references break if this isn't loaded?)
\usepackage{mdframed} % to easily handle boxes
\usepackage[misc]{ifsym} % use \Letter symbol for corresponding author

% customization
\setcounter{Maxaffil}{0} % affiliations
\renewcommand\Affilfont{\itshape\small} % style of affiliations text
\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
\addto{\captionsenglish}{\renewcommand{\bibname}{References}}
\newcommand{\idea}[1]{\textcolor{red}{#1}}
\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
\newcommand{\Burl}[1]{\url{#1}}
\zxrsetup{toltxlabel} % cross references: draft & suppl mat
\zexternaldocument*[M-]{draft}
\hypersetup{
  colorlinks = true,
  citecolor=  black,
  linkcolor = {blue},
  filecolor = cyan % controls color of external ref, if used
}
\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Figure}} % custom figure captions
\captionsetup[table]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Table}} % custom table captions
\hypersetup{citecolor={blue}}
\definecolor{lightgray}{gray}{0.9}
\newcommand{\figref}[2][]{%
  \hyperref[{#2}]{%
    Figure~\ref*{#2}%
    \ifx\\#1\\%
    \else
      #1%
    \fi
  }%
}
\newcommand{\tableref}[1]{%
  \hyperref[{#1}]{%
   Table~\ref*{#1}%
  }%
}
\newcommand{\methodsref}[1]{%
  \hyperref[{methods:#1}]{%
   Methods:~\nameref*{methods:#1}%
  }%
}

@@

\affil[$\textrm{\Letter}$]{\normalfont~alvaro.sanchez@yale.edu}
\affil[*]{\normalfont~These authors contributed equally}
\date{}

@@

Microbial communities often invade one another.
This has been observed, for instance, 
in river courses where terrestrial microbes mix with aquatic microorganisms
\cite{Mansour2018,Luo2020,Vass2021}
or in soil communities being invaded as a result of
tillage and outplanting \cite{Rillig2016} or by
aerially dispersed bacteria and funghi \cite{Evans2019}.
Gut microbiomes can invade external communities
through the host animal secretions \cite{Dutton2021},
and the skin microbiota is also subject to
invasions when making contact with environmental sources of microbes \cite{Vandegrift2019}.

@@

The mechanisms of ecological co-selection during community coalescence are still poorly understood.
Do a few key species recruit everyone else, or are collective interactions among all species
(including the rarer members of the community) relevant for coalescence outcomes?
While it is reasonable to expect species with larger population sizes to have a proportionally oversized effect,
natural communities tend to be highly diverse \cite{Louca2016}
and
the role played by the less abundant species has long been subject to debate \cite{Winfree2015}.
Laboratory cultures have also been found to contain uneven distributions of multiple strains that feed
off the metabolic secretions of the dominant species \cite{Rosenzweig1994,Goldford2018}.
The fate of these sub-dominant taxa may be dependent on the invasional success of their dominant
species, or, alternatively, the dominant itself may owe its dominance (at least in part) to cross-feeding
or other forms of facilitation from the rarer members of the population.
These scenarios would give rise to ``top-down'' or ``bottom-up'' community cohesiveness, respectively.
Either of these forms of co-selection could, in principle, be positive (recruitment) or negative (antagonism),
as illustrated in \figref[e]{fig1}.
Which of these situations are typically found in nature?
Previous theoretical and computational studies suggest that the answer is determined by the type and
strength of the interactions of the community members with one another and with the environment
\cite{Tikhonov2016,Vila2019,Lechon2021},
but addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.

In previous work, we have shown that a large amount of soil and plant microbiomes can be cultured
\textit{ex situ} in synthetic minimal environments with a single supplied limiting resource under serial
growth-dilution cycles \cite{Goldford2018} (\figref[a-b]{fig1}). Under these conditions, environmental
microbiomes spontaneously re-assemble into complex multi-species communities sustained by dense
cross-feeding facilitation networks \cite{Goldford2018}. In addition,
and just like in natural consortia,
species abundance distributions
in these communities are generally long-tailed and uneven (\figref[d]{fig1} and \figref{figS1}),
with the dominant (most abundant) species typically comprising most of the biomass ($\text{median}=46\%$,
\figref{figS1}). Because these communities are easy to manipulate and grow in high throughput, and
are largely made up by culturable members, they represent good test cases to investigate ecological
co-selection during community coalescence. Here we focus on the dominants and ask whether they can
co-select or be co-selected by the sub-dominant species in their communities (henceforth referred to as
their \textit{cohorts}, \figref[c]{fig1}).

Our results indicate that co-selection varies in direction and strength depending on the supplied limiting
resource.
This primary resource, in turn, has been shown to
shape the structure and interactions of the communities \cite{Estrela2020}.
We observe that, when top-down co-selection is weak, bottom-up co-selection can be very strong,
with positive co-selection being far more common than negative co-selection.
\iffalse
Our results confirm that positive top-down co-selection is common, but its effects are weak.
In contrast, bottom-up co-selection can be very strong, and positive co-selection is far more common
than negative co-selection.
\fi
We then turn to a Microbial Consumer-Resource Model (MicroCRM)
\cite{Goldford2018,Marsland2019,Marsland2020}
that is able to capture the dynamics of microbial communities dominated by metabolic interactions,
as is the case for the ones assembled in our experimental conditions \cite{Goldford2018,Estrela2020}.
We show that the empirically observed trends in ecological co-selection are reproduced
with minimal model assumptions, and that the recurrence of top-down and bottom-up co-selection
is determined by the configuration of the cross-feeding networks in the MicroCRM.
Our findings indicate that collective interactions play an important role at dictating community structure
during coalescence.

@@

We collected eight natural microbiomes from different soil and plant environmental samples
(\figref[a]{fig1})
and used them to inoculate our synthetic communities, which were stabilized 
in serial batch-culture biorreactors for 84 generations in synthetic minimal media containing either
glutamine or citrate as the only supplied carbon source
(\figref[b]{fig1}, \methodsref{community-assembly}).
We chose these two carbon sources because they are metabolized through different
pathways in bacteria \cite{Dimroth2004,Forchhammer2007},
and we hypothesize that communities assembled in either resource will be supported by
cross-feeding networks of distinct sets of metabolites \cite{Goldford2018,Estrela2020}
thus leading to potentially variable degrees of community cohesiveness
and coalescence outcomes \cite{Tikhonov2016,Tikhonov2017,Castledine2020,Lechon2021}.
We isolated the dominant species of every community
(\methodsref{dominants})
and identified them by Sanger-sequencing their 16S rRNA gene
(\methodsref{sequencing}), which correctly matched the dominant
Exact Sequence Variant (ESV) \cite{Callahan2016,Callahan2017} found through
community-level 16S Illumina sequencing
(\figref{figS1}).
These dominants remained at high frequency after seven additional transfers
with the exception of two of the citrate communities and one of the glutamine
communities (where the dominants were presumably a transiently dominating species)
that were excluded from further analysis
(\figref{figS1}).
Similarly, pairs of communities where the dominants shared a same 16S sequence
and had similar colony morphology were excluded
(\figref{figS1}).

@@

If communities being coalesced were highly cohesive from the top-down,
the dominant species would co-select the rarer members of its community during coalescence
(\figref[e]{fig1}, left panels).
In this scenario, we would expect the outcome of community coalescence to be predicted by
which of the two dominants is most competitive in pairwise competition.
Analogously, competition between dominants should be affected only weakly by the presence or
absence of the cohorts, that would play a passive role under these conditions.
To test this hypothesis, we performed all pairwise competitions between dominant species
in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
the cultures for seven serial transfers, roughly 42 generations
(\methodsref{competitions}).
We then carried out all possible pairwise community coalescence experiments by mixing
equal volumes of the communities and propagating the resulting cultures for seven extra
transfers
(\figref[f]{fig1}).
\iffalse
and we quantified coalescence outcomes using the similarity between the coalesced and the
invasive communities
(\methodsref{metrics}).
\fi
The frequencies of all species in both community-community and dominant-dominant competitions
were determined by 16S Illumina sequencing
(\methodsref{sequencing}).

We found that, for communities assembled in the glutamine environment,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. during community coalescence
(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
This correlation is significantly higher for the citrate communities
 (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).
This suggests that, in the glutamine environments, head-to-head competition of dominants
is heavily influenced by higher order effects introduced by the rare taxa of the communities.
On the other hand, the cohorts seem to play a more passive role in the citrate environments.
To test the effects of top-down co-selection at the community level,
we quantified the distances between the invasive and coalesced communities
using the relative Bray-Curtis similarity (\methodsref{metrics})
and compared them to the outcomes of the pairwise competitions between dominants alone.
We again noticed differences between glutamine and citrate communities:
for the former,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(\figref[b]{fig2} left panel, $R^2=0.15$, $p<0.05$).
For the latter,
the structure of the coalesced communities tends to be more strongly dictated by the
result of the dominant-dominant competition
(\figref[b]{fig2} middle panel, $R^2=0.57$, $p<10^{-4}$).
Alternative quantifications of community distance yield similar results,
with weaker effects when the metric used
accounts only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).

Together, these observations suggest that the strength of top-down co-selection depends on
the environment where communities are assembled and coalescence takes place.
Communities stabilized with citrate as the primary supplied resource
display a strong degree of top-down cohesiveness,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to just be passively responding to their dominants
but rather playing an active role in community coalescence.

To investigate the determinants of top-down co-selection and the factors modulating its strength,
we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
Model (MicroCRM) \cite{Goldford2018,Marsland2019} as implemented in the Community Simulator package
for Python \cite{Marsland2020}
(\hyperref[box1]{Box~1}).
We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020}
for which the MicroCRM provides a good description.
Based on previous experimental work
\cite{Harcombe2014,Pinu2018,Estrela2020},
we allowed different species in our model
to secrete different sets of byproducts when metabolizing a same resource
(\hyperref[box1]{Box~1}).
\iffalse
this choice is discussed in \hyperref[box1]{Box~1}.
\fi
To reproduce our experimental protocol \textit{in silico},
we first generated a library of resources and two non-overlapping pools of species.
Each pool was used to seed a collection of 100 invasive
and 100 resident communities respectively
by randomly choosing 50 species and
allowing them to stabilize through 20 growth-dilution cycles.
We then mixed these stable communities in pairs to simulate our coalescence
and dominant-dominant competition experiments
(\methodsref{sim}).
We found that the MicroCRM was able to capture a correlation between
the head-to-head pairwise competition of dominants
and the outcome of community coalescence
(\figref[b]{fig2}, right panel),
further supporting the idea that top-down ecological co-selection can consistently
emerge from metabolic interactions across species.

\iffalse
[Maybe this ``outro'' sentence is not even necessary]
Our experiments had shown that this correlation varied
across communities assembled in the glutamine and citrate environments
(\figref[b]{fig2}, left and middle panels),
which led us to investigate the determinants of the strength and direction
of ecological co-selection in our model.

\fi
\clearpage

@@

The Microbial Consumer-Resource Model (MicroCRM) \cite{Goldford2018,Marsland2019,Marsland2020}
is a modeling framework based on the classic MacArthur's consumer resource model
\cite{MacArthur1970}.
It encodes the dynamics of a system with $S$ species and $M$ resources
in terms of a consumer preference matrix $\mathbf{c}$ and a metabolic matrix $\mathbf{D}$,
with an additional set of parameters controlling
the species maintenance costs ($m_i$ for species $i$),
the resource energy densities ($w_\alpha$ for resource $\alpha$),
the energy to growth rate conversion factor ($g_i$ for species $i$)
and the leakage fraction, i.e. the amount of energy lost as byproducts when a resource is consumed
($l_\alpha$ for resource $\alpha$).
The element $c_{i\alpha}$ of the consumer preference matrix
represents the uptake rate of resource $\alpha$ by
species $i$ (although the relationship between $c_{i\alpha}$ and the uptake rate can be more
complex in modeling scenarios that are not considered here,
see \cite{Goldford2018,Marsland2019,Marsland2020}).
Experimental evidence suggests that individual species can secrete different sets of metabolites
to the environment when growing on a same primary resource
\cite{Harcombe2014,Pinu2018,Estrela2020}.
Thus, we define $\mathbf{D}$ as a three-dimensional matrix
where the element $D_{i\beta\alpha}$ represents the energy flux in the form of resource $\beta$
that is secreted by species $i$ when it metabolizes resource $\alpha$.
Note that $D_{i\beta\alpha}$ need not be equal to $D_{j\beta\alpha}$ if $i \neq j$
(see illustration below).
\iffalse
For our simulations,
the elements of $\mathbf{c}$ are sampled from a gamma distibution,
while the elements of $\mathbf{D}$ are sampled from a Dirichlet distribution
to ensure that the normalization condition $\sum_\beta D_{i\beta\alpha} = 1$
is satisfied for every $i$ and $\alpha$.
\fi
\iffalse
The sampling is performed so that species belonging to a same family
tend to secrete more similar byproducts when consuming a same resource
(see \methodsref{sim} and \cite{Marsland2019,Marsland2020}),
as supported by experimental observations
\cite{Harcombe2014,Pinu2018,Estrela2020}.
\fi

@@

\iffalse
We found that, for communities assembled in the glutamine environment,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
but this correlation was stronger for the citrate communities
(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
Alternative quantifications of community distance yielded similar results,
with weaker effects when the metric used
accounted only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).
Furthermore, we observed that, in the glutamine communities,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. after coalescence
(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
On the other hand, pairwise competition between dominants of the citrate communities
is highly correlated with the relative abundance of said dominants in community coalescence
 (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).

Together, these results suggest that the strength of top-down co-selection depends on
the primary resource supplied to the coalesced communities.
Communities assembled with citrate as the primary supplied resource
display a strong degree of top-down co-selection,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to be passively responding to their dominants.
Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
This finding led us to investigate the potential role of bottom-up ecological co-selection
(\figref[e]{fig1}, right panels), i.e. whether in some cases the dominants may be co-selected
for or against by their cohorts.
\fi

@@

Our data indicates that the primary resource supplied to the communities can modulate
the effect that the cohorts have in the dominants pairwise competition
(\figref[a]{fig2})
and the strength of top-down co-selection
(\figref[b]{fig2}, left and middle panels).
The fact that
our model captures these trends
suggests that this might be a result of the metabolic interactions
between community members, including the rarer taxa.
To investigate the potential role of the cohorts in coalescence,
 i.e. whether the dominants may be co-selected for or against by them
(\figref[e]{fig1}, right panels),
we ran a new set of simulations
this time invading resident communities with the dominants alone
(\methodsref{sim}).
We compared the invasion success of the dominants in isolation with respect to our previous
simulations where they invaded accompanied by their cohorts.
The invasion success of the dominants was quantified by their relative abundance
in the final stabilized communities.
Whenever positive bottom-up ecological co-selection is strong, we expect to see dominants
reaching higher invasion success with their cohorts than by themselves
(\figref[b]{fig3}, green shaded region).
On the other hand, a high degree of bottom-up antagonism would result in dominants
invading more effectively alone than in the presence of their cohorts
(\figref[b]{fig3}, red shaded region).
Alternatively, if both forms of bottom-up co-selection are weak, we would see a similar invasion success
regardless of the presence or absence of the cohort
(\figref[b]{fig3}, gray shaded region).

>>>>

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{upgreek} % upright greek letters for math
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{makecell} % more table customization
\usepackage{mathtools} % nice matrices
\usepackage{units} % nice fractions
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[table,xcdraw]{xcolor} % text and table background colors
\usepackage[margin=2.5cm]{geometry} % margins
\usepackage{lineno} % line numbers
\usepackage{textcomp} % special characters (some references break if this isn't loaded?)
\usepackage{mdframed} % to easily handle boxes
\usepackage[misc]{ifsym} % use \Letter symbol for corresponding author

% customization
\setcounter{Maxaffil}{0} % affiliations
\renewcommand\Affilfont{\itshape\small} % style of affiliations text
\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in references
\addto{\captionsenglish}{\renewcommand{\bibname}{References}}
\newcommand{\idea}[1]{\textcolor{red}{#1}}
\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
\newcommand{\Burl}[1]{\url{#1}}
\hypersetup{
  colorlinks = true,
  citecolor=  black,
  linkcolor = {blue},
  filecolor = cyan % controls color of external ref, if used
}
\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Figure}} % custom figure captions
\hypersetup{citecolor={blue}}
\definecolor{lightgray}{gray}{0.9}
\newcommand{\figref}[2][]{%
  \hyperref[{#2}]{%
    Figure~\ref*{#2}%
    \ifx\\#1\\%
    \else
      #1%
    \fi
  }%
}
\newcommand{\tableref}[1]{%
  \hyperref[{#1}]{%
   Table~\ref*{#1}%
  }%
}
\newcommand{\methodsref}[1]{%
  \hyperref[{methods:#1}]{%
   Methods:~\nameref*{methods:#1}%
  }%
}

@@

\affil[$\textrm{\Letter}$]{\normalfont alvaro.sanchez@yale.edu}
\affil[*]{\normalfont These authors contributed equally}
\date{}

@@

Microbial communities often invade one another.
This has been observed, for instance, 
in river courses where terrestrial microbes mix with aquatic microorganisms,
\cite{Mansour2018,Luo2020,Vass2021}
or in soil communities being invaded as a result of
tillage and outplanting \cite{Rillig2016} or by
aerially dispersed bacteria and funghi \cite{Evans2019}.
Gut microbiomes can invade external communities
through the host's secretions \cite{Dutton2021},
and the skin microbiota is also subject to
invasions when it makes contact with environmental sources of microbes \cite{Vandegrift2019}.

@@

The mechanisms of ecological co-selection during community coalescence are still poorly understood.
Do a few key species recruit everyone else, or are collective interactions among all species
(including the rarer members of the community) relevant for coalescence outcomes?
While it is reasonable to expect species with larger population sizes to have a proportionally oversized effect,
natural communities tend to be highly diverse \cite{Louca2016}
and
the role played by the less abundant community members has long been subject to debate \cite{Winfree2015}.
Laboratory cultures have also been found to contain uneven distributions of multiple strains that feed
off the metabolic secretions of the dominant species \cite{Rosenzweig1994,Goldford2018}.
The fate of these sub-dominant taxa may be dependent on the invasion success of their dominant
species, or, alternatively, the dominant itself may owe its dominance (at least in part) to cross-feeding
or other forms of facilitation from the rarer members of the population.
We refer to these two opposite scenarios as
``top-down'' or ``bottom-up'' forms of community cohesiveness, respectively.
Top-down cohesiveness emerges when the dominant invader co-selects other sub-dominant taxa
into the final community during coalescence.
Alternatively, bottom-up cohesiveness refers to the case when the dominant is co-selected by
the more rare members of its community.
Either of these forms of co-selection could, in principle, be positive (recruitment) or negative (antagonism),
as illustrated in \figref[e]{fig1}.
Which of these situations are typically found in nature?
Previous theoretical and computational studies suggest that the answer is determined by the type and
strength of the interactions of the community members with one another and with the environment
\cite{Tikhonov2016,Vila2019,Lechon2021},
but addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.

In previous work, we have shown that a large amount of soil and plant microbiomes can be cultured
\textit{ex situ} in synthetic minimal environments with a single supplied limiting resource under serial
growth-dilution cycles \cite{Goldford2018} (\figref[a-b]{fig1}). Under these conditions, environmental
microbiomes spontaneously re-assemble into complex multi-species communities sustained by dense
cross-feeding facilitation networks \cite{Goldford2018}. In addition,
and just like in natural consortia,
species abundance distributions
in these communities are generally long-tailed and uneven (\figref[d]{fig1} and \figref{figS1}),
with the dominant (most abundant) species typically comprising most of the biomass ($\text{median}=46\%$,
\figref{figS1}). Because these communities are easy to manipulate and grow in high throughput,
they represent good test cases to investigate ecological
co-selection during community coalescence. Here we focus on the dominants and ask whether they can
co-select or be co-selected by the sub-dominant species in their communities (henceforth referred to as
their \textit{cohorts}, \figref[c]{fig1}).

Our results indicate that
when top-down co-selection is weak, bottom-up co-selection can be very strong,
with positive co-selection being far more common than negative co-selection.
We then turn to a Microbial Consumer-Resource Model (MicroCRM)
\cite{Goldford2018,Marsland2019,Marsland2020}
that is able to capture the dynamics of microbial communities dominated by metabolic interactions,
as is the case for the ones assembled in our experimental conditions \cite{Goldford2018,Estrela2020}.
We show that the empirically observed trends in ecological co-selection are reproduced
with minimal model assumptions, and that the recurrence of top-down and bottom-up co-selection
is determined by the configuration of the cross-feeding networks in the MicroCRM.
Our findings indicate that collective interactions play an important role at dictating community structure
during coalescence.

@@

We collected eight natural microbiomes from different soil and plant environmental samples
(\figref[a]{fig1})
and used them to inoculate
eight identical habitats containing minimal media with either glutamine or citrate as
the only supplied carbon source.
We chose these two carbon sources because they are metabolized through different
pathways in bacteria \cite{Dimroth2004,Forchhammer2007},
and we hypothesize that communities assembled in either resource will be supported by
cross-feeding networks of distinct sets of metabolites \cite{Goldford2018,Estrela2020},
thus leading to potentially variable degrees of community cohesiveness
and coalescence outcomes \cite{Tikhonov2016,Tikhonov2017,Castledine2020,Lechon2021}.
After inoculation, all communities were serially passaged for 12 transfers (84 generations),
with an incubation time of 48 hours and a dilution factor of 1:100.
(\figref[b]{fig1}, \methodsref{community-assembly}).
In previous work we have shown that under these conditions,
12 transfers allow communities to approach a state of ``generational equilibrium'',
where the community composition at the end of one batch incubation will be the same as in consecutive incubations.
We isolated the dominant species of every community
(\methodsref{dominants})
and identified them by Sanger-sequencing their 16S rRNA gene
(\methodsref{sequencing}), which correctly matched the dominant
Exact Sequence Variant (ESV) \cite{Callahan2016,Callahan2017} found through
community-level 16S Illumina sequencing
(\figref{figS1}).
These dominants remained at high frequency after seven additional transfers
with the exception of two of the citrate communities and one of the glutamine
communities (where the dominants were presumably a transiently dominating species)
that were excluded from further analysis
(\figref{figS1}).
Similarly, pairs of communities where the dominants shared a same 16S sequence
and had similar colony morphology were excluded
(\figref{figS1}).

@@

One form of cohesiveness may arise when the sub-dominant members of the community
depend on the dominant species.
This can occur, for instance, when the dominant provides resources (or stressors)
that select for the sub-dominant taxa
(\figref[e]{fig1}, left panels).
If communities being coalesced are highly cohesive from the top-down,
the fate of the sub-dominant community members will be tied to their dominant:
if it gets excluded, they will be likely to fall with it, and if it is able to resist coalescence,
they will be likely to follow suit.
In this scenario, we would expect the outcome of community coalescence to be predicted by
which of the two dominants is most competitive in pairwise competition.
Likewise, competition between dominants should be affected only weakly by the presence or
absence of sub-dominant species, which would play a passive role under these conditions.
To test this hypothesis, we performed all pairwise competitions between dominant species
in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
the cultures for seven serial transfers, roughly 42 generations
(\methodsref{competitions}).
We then carried out all possible pairwise community coalescence experiments by mixing
equal volumes of the communities and propagating the resulting cultures for seven extra
transfers
(\figref[f]{fig1}).
The frequencies of all species in both community-community and dominant-dominant competitions
were determined by 16S Illumina sequencing
(\methodsref{sequencing}).

To test the effects of top-down co-selection at the community level,
we quantified the distances between the invasive and coalesced communities
using the relative Bray-Curtis similarity (\methodsref{metrics})
and compared them to the outcomes of the pairwise competitions between dominants alone.
We noticed a difference between communities assembled in the glutamine and citrate environments:
for the latter,
the structure of the coalesced communities tends to be strongly dictated by the
result of the dominant-dominant competition
(\figref[a]{fig2} middle panel, $R^2=0.57$, $p<10^{-4}$).
For the former,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(\figref[a]{fig2} left panel, $R^2=0.15$, $p<0.05$).
Alternative quantifications of community distance yield similar results,
with weaker effects when the metric used
accounts only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).
We then examined whether, as predicted by the top-down cohesiveness hypothesis,
the cohorts would play a passive role on the competition between dominant species.
We found that, for communities assembled in the citrate environments,
the relative frequency of a dominant against another in head-to-head pairwise competition
is highly predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. during community coalescence
(\figref[b]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).
This is not the case for the glutamine communities
(\figref[b]{fig2} red dots, $R^2=0.04$, $p>0.05$).
This suggests that, in the glutamine environments, head-to-head competition of dominants
is heavily influenced by interactions between those dominants and the rarer taxa of the communities.
On the other hand, the cohorts seem to play a more passive role in the citrate environments.
Together, these observations indicate that
communities stabilized with citrate as the primary supplied resource
display a strong degree of top-down cohesiveness,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to just be passively responding to their dominants
but rather playing an active role in community coalescence.

To investigate the determinants of top-down co-selection and the factors modulating its strength,
we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
Model (MicroCRM) \cite{Goldford2018,Marsland2019} as implemented in the Community Simulator package
for Python \cite{Marsland2020}
(\hyperref[box1]{Box~1}).
We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multi-species communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020}
for which the MicroCRM provides a good description.
We and others have previously found a strong concordance between the behavior of
laboratory and natural microbial communities and the generic behavior of the MicroCRM
\cite{Goldford2018,Marsland2019,Marsland2020,Marsland2020b,Estrela2021}.
To reproduce our experimental protocol \textit{in silico},
we first generated a library of resources and two non-overlapping pools of species.
Each pool was used to seed a collection of 100 invasive
and 100 resident communities respectively
by randomly choosing 50 species and
allowing them to stabilize through 20 growth-dilution cycles.
We then mixed these stable communities in pairs to simulate our coalescence
and dominant-dominant competition experiments
(\methodsref{sim}).
We found that the MicroCRM simulations naturally exhibit the observed correlation between
the head-to-head pairwise competition of dominants
and the outcome of community coalescence
(\figref[a]{fig2}, right panel),
further supporting the idea that top-down ecological co-selection consistently
emerges from metabolic interactions across species.
Moreover, we found that top-down co-selection is observed under a wide range of different simulation conditions
and cross-feeding networks (\figref{figSX}), indicating that it is a robust phenomenon.

@@

The Microbial Consumer-Resource Model (MicroCRM) \cite{Goldford2018,Marsland2019,Marsland2020}
is a modeling framework based on the classic MacArthur's consumer resource model
\cite{MacArthur1970}.
It encodes the dynamics of a system with $S$ species and $M$ resources
in terms of a consumer preference matrix $\mathbf{c}$ and a metabolic matrix $\mathbf{D}$,
with an additional set of parameters controlling
the species maintenance costs ($m_i$ for species $i$),
the resource energy densities ($w_\alpha$ for resource $\alpha$),
the energy to growth rate conversion factor ($g_i$ for species $i$)
and the leakage fraction, i.e. the amount of energy lost as byproducts when a resource is consumed
($l_\alpha$ for resource $\alpha$).
The element $c_{i\alpha}$ of the consumer preference matrix
represents the uptake rate of resource $\alpha$ by
species $i$ (although the relationship between $c_{i\alpha}$ and the uptake rate can be more
complex in modeling scenarios that are not considered here,
see \cite{Goldford2018,Marsland2019,Marsland2020}).
Experimental evidence suggests that individual species can secrete different sets of metabolites
to the environment when growing on a same primary resource
\cite{Harcombe2014,Pinu2018,Estrela2020}.
Thus, we define $\mathbf{D}$ as a three-dimensional matrix
where the element $D_{i\beta\alpha}$ represents the energy flux in the form of resource $\beta$
that is secreted by species $i$ when it metabolizes resource $\alpha$.
Note that $D_{i\beta\alpha}$ need not be equal to $D_{j\beta\alpha}$ if $i \neq j$
(see illustration below).

@@

@@

Our data indicates that the primary resource supplied to the communities can modulate
the effect that the cohorts have in the dominants pairwise competition
(\figref[b]{fig2})
and the strength of top-down co-selection
(\figref[a]{fig2}, left and middle panels).
The fact that
our model captures these trends
suggests that this might be a result of the metabolic interactions
between community members, including the rarer taxa.
To investigate the potential role of the cohorts in coalescence,
 i.e. whether the dominants may be co-selected for or against by them
(\figref[e]{fig1}, right panels),
we ran a new set of simulations
this time invading resident communities with the dominants alone
(\methodsref{sim}).
We compared the invasion success of the dominants in isolation with respect to our previous
simulations where they invaded accompanied by their cohorts.
The invasion success of the dominants was quantified by their relative abundance
in the final stabilized communities.
Whenever positive bottom-up ecological co-selection is strong, we expect to see dominants
reaching higher invasion success with their cohorts than by themselves,
with the strongest instances occurring when dominants are unable to invade on their own
but reach high densities when invading together with their cohorts
(\figref[b]{fig3}, green shaded region).
Alternatively, a high degree of bottom-up antagonism would result in dominants
invading more effectively alone than in the presence of their cohorts
(\figref[b]{fig3}, red shaded region).
Finally, if bottom-up co-selection is weak, we would see a similar invasion success
regardless of the presence or absence of the cohort
(\figref[b]{fig3}, gray shaded region).
